---
title: "Filling Missing Data in Climatological Time-Series"
author: '  Manon von Kaenel & Nick Depsky  '
date: '  July 7th 2016  '
output:
  html_document:
    number_sections: no
    theme: readable
    toc: yes
  pdf_document:
    toc: yes
---

Some Notes for filling missing data in climatological time-series using R :cloud: :cyclone: :sunny:
SEI R-users group 7-7-2016

The Rmarkdown file and data used to generate this page is stored on [github](https://github.com/njdepsky/RSpatial_6-29)

## Goals 
Go over some common workflows relevant to handling missing data when dealing with climatological station time-series data (This example will use precipitation records from the Santa Clara Valley, CA). In particular...  

* Importing Data and Calculating Distance between stations
* Calculating Temporal Record Overlap and Correlation Coefficients between stations

## Resources
* Data used for the examples are hosted in this repository. The zipped file contains the following:
    - `SCVWD_Daily_RainGage_ts.csv` : csv of the station precipitation data in Santa Clara County, CA
    - `Station_XY_ft.csv`: csv of station Lon Lat Coordinates
You can download/unzip the data by hand or run the following code to download and unzip the data to a temporary directory

```{r eval = FALSE}
setwd( tempdir() )  # set working directory to temporary directory 
download.file("https://github.com/njdepsky/Climate-Data-Processing-for-WEAP-Applications/raw/master/rawdata/fillmissdata.zip", destfile = 'fillmissdata.zip')
unzip( 'fillmissdata.zip' , exdir = '.')
```

# 0. Prerequisites
Some date functions we will use come from the `zoo` package, so be sure to have it installed prior to attempting this exercise. You can download this package with...

```{r eval=FALSE}
install.packages('zoo', repos = "http://cran.cnr.berkeley.edu/")
```

# 1. Get the data in

```{r}
library(zoo)
```

## Loading a CSV file of Precipitation Station Records
```{r}
rain <- read.csv('SCVWD_Daily_RainGage_ts.csv', check.names = F, stringsAsFactors = F)
```
that was easy!

as a side note, the 'check.names = F' parameter in the 'read.csv' function preserves the column headers from the original record table
